\documentstyle[12pt]{report}
\pagestyle{headings}

\begin{document}

\pagestyle{empty}

\title{Dominion Gamemaster's Manual}

\author{Mark Galassi}

\date{\today}

\maketitle

\pagestyle{headings}

\tableofcontents

\chapter{Overview}
This document gives some hint to the Game Master on how to run a game
of {\bf Dominion}.  It is {\em extremely} incomplete.

\chapter{Making a world}

Look at the README file to see how to compile dominion.

Once you have run ``make new-world'', everything should be set up.
You might want to make sure that the ownership of the files in LIBDIR
is correct, and that the prgram {\em dominion} is setuid to that
owner (using the {\em chmod} command).

The world is made with one nation already in it.  This nation is
called {\em Gamemaster}, of race {\em Creator} and magic order {\em
Master}.  The Gamemaster nation is not really to be played as a
nation, since it does not have any sectors, or any other desireable
properties.  But it has the most totally bodacious [E] command which
will be described below.

\chapter{Adding a nation}

You run the program dom\_add.  You have to run this with the userid of
the owner of the game.  You will be prompted interactively for all the
paramters of this nation.  Then the nation will be added to the world
and they can play.  Nations should be added when nobody is playing,
and while you are adding anyone in the game will be kicked out.

\chapter{News and mail}

You will be prompted to set up some newsgroups for the game.  There is
always a newsgroup called {\em News} in which general information is
posted about the update, battles, and so on.  You should create
another newsgroup which can be posted to by all nations.  This gives
the players a public forum which can induce some great role-playing.

Mail is sent from one player to the other within the game.  A player
can set a {\em forwarding address} with the [O]ptions menu.  The mail
will be sent to the electronic mail address they specify.

News can also be forwarded: the Gamemaster creates a file called
``news.post'' which contains a list of electronic mail addresses.
This file should reside in the main dominion LIBDIR.  Any news article
will be mailed to everyone in that list when it gets posted.

\chapter{Special Gamemaster commands}

The Gamemaster can do special things to the game in two ways: by {\bf
editing the ``exec files''} and by using the {\bf Edit Sector} and {\bf
Edit Nation} commands.

Editing the ``exec files'' requires some knowledge of how the game
works internally, though some people who have not seen the source code
have been able to work on this.

The [E]dit command to edit a [s]ector or a [n]ation is menu-driven, so
you can experiment with it.  It allows you to change their storage of
food, metal, jewels, spell points and so on.  It also allows you to
change the population of a sector, or to add to a nation's magic and
technological skill.

You should be careful using these Game Master commands, since some of
them have not yet been tested in all possible contexts.  You should be
most careful about changing the ownership of a sector.  For now it has
been tested only for changing sector ownership from {\em unowned} to
being owned by a given nation.  Changing ownership from one nation to
another has not been tested.  In fact, I suspect it will not work.

\chapter{Using the Master race and order}

The magic order Master can be used to create special nations run by
the Game Master, since it gets all spells and spirits for free.  This
way you can do some active ``Gamemaster'' play, if you so wish.

You should not give the magic order Master to any ordinary player.

more here....

\end{document}
